\large \textbf{Title:} The role of learning in the evolution of status signalling: a modeling approach

### 2023-09-14 manuscript/manuscript_1.0.tex
Title_page rmd file and readme md file, getting readu for realease and submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdfauthor={Andrés E. Quiñones; C. Daniel Cadena*; Redouan Bshary},
  pdfkeywords={My keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \usepackage{float}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
 \newcommand{\beginsupplement}{ \setcounter{table}{0}     \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Learning can mediate both badges of status and individual
recognition}
\author{Andrés E. Quiñones\footnote{Laboratorio de Biología Evolutiva de
  Vertebrados, Departamento de Ciencias Biológicas, Universidad de los
  Andes, Bogotá, Colombia} \and C. Daniel Cadena* \and Redouan
Bshary\footnote{Institute of Biology, University of Neuchâtel,
  Neuchâtel, Switzerland}}
\date{July 18, 2022}

\begin{document}
\maketitle
\begin{abstract}
Adaptive behavioural responses often depend on qualities of the
interacting partner of an individual. For example, when competing for
resources, an individual might be better off escalating fights with
individuals of lower quality, while restraining from fighting
individuals of higher quality. Communication systems involving signals
of quality allow individuals to reduce uncertainty regarding the
fighting ability of their partners and make more adaptive behavioral
decisions. However, dishonest individuals can destabilize such
communications systems. One open question is whether cognitive
mechanisms, such as learning, can maintain the honesty of signals, thus,
favoring their evolutionary stability. Here we present evolutionary
simulations, where individuals can produce a signal proportional to
their quality; and learn along their lifetime the best response to the
signal emitted by their peers. In these simulations, learning on the
receiver side can mediate the evolution of signals of quality on the
sender side. When the cost of the signal is proportional to the quality
of the sender populations are only composed of honest signalers. When
the cost is not proportional to the quality of the signaler, the
population is composed of both honest and dishonest signalers. We argue
that learning can be a general cognitive mechanism playing a role in a
wide range of communication systems.
\end{abstract}

@@

The outcome of interactions with con-specifics is an important
determinant of fitness in social animals. Irrespective of whether such
interactions are cooperative or competitive, the actions of interacting
individuals partly determine their reproductive success. However, the
best action for an individual might vary depending on its condition, and
that of individuals with whom it interacts. Thus, acting based on
information about interacting partners is typically adaptive (Quiñones
et al. 2016). Acquiring this information, however, is far from trivial.
In some cases, interacting partners (signallers) might be `willing' to
provide accurate information, but in others it might be in their own
interest to conceal information (Johnstone 1997), or to provide
erroneous information (Johnstone and Norris 1993). Typically, in a given
context, some proportion of individuals will benefit from broadcasting
accurate information, whereas others will benefit from hiding it. Take,
for instance, an interaction between two individuals where one can help
the other. Given some costs and benefits, a potential donor will be
interested in helping related individuals. Therefore, for a relative of
the donor, broadcasting kinship is advantageous, whereas for an
unrelated individual, concealing the lack of kinship is better. Similar
scenarios may apply to many aspects of social life such as finding
mates, feeding offspring, dominance relationships and aggressive
interactions (Bradbury and Vehrencamp 2011; Møller 1988; Tibbetts and
Dale 2004). In dominance and aggressive interactions, an important piece
of information to guide individual actions is interacting partners
fighting ability, which is sometimes referred to as Resource Holding
Potential (RHP) (Parker 1974), or simply as quality. Responsiveness to
the quality of the partner is central to communication systems such as
badges of status, where an arbitrary signal conveys quality (Johnstone
and Norris 1993; Rohwer 1975). Here, arbitrary refers to a signal which
is not ontogenetically correlated with quality. Badges of status can be
evolutionarily stable whenever the signal imposes a fitness cost which
decreases with the quality of an individual (Botero et al. 2010;
Johnstone and Norris 1993), such that it's no longer in the interest of
low quality individuals to fake quality by producing a large badge. The
cost of honest signals is a more general principle of communications
systems, usually referred to as the handicap principle (Grafen 1990;
Zahavi 1975).

A somewhat ignored component of communication systems, in the context of
aggressive interactions, is the cognitive aspects of the receiver
module. Theoretical models often assume that Communication systems
relying on badges of status to have reaction norms as their mechanistic
underpinning (Botero et al. 2010), with individuals using their own
quality and the opponent's badge to determine whether to aggressively
engage in contest. Reaction norms allow individuals to respond to the
available information without a big cognitive burden. That is at least
in contrast to other information processing mechanisms like individual
recognition. With individual recognition individuals associate cue of
their peers to their quality. These association must be learn thoughout
individuals lifes, thus it has the usual cognitive requirements of an
associative learning process. An alternative view of badges of status is
that individuals learn to react to them based on their experiences
(Guilford and Dawkins 1991), which would imply that receivers must learn
to associate signals with the fighting ability of bearers just as in
systems based on individual recognition. In cases where badges of status
vary quantitatively (e.g.~in size or intensity), fighting ability may
increase monotonically with attributes of the badge and would be
reinforced by every interaction, so in principle learning would be
faster than in systems involving individual recognition in which the
association between signals and their meaning would vary depending on
the interacting partners. In any case, learning could be a central
cognitive mechanism in both types of communication systems, but the role
of learning in these contexts has not been thoroughly explored in the
empirical nor theoretical literature.

Associative learning is a key cognitive mechanism that allows
individuals to associate rewards with environmental stimuli and thus
behave adaptively (Staddon 2016). Associative learning exists in all
major vertebrate taxa, and in many invertebrates as well (Heyes 2012;
Macphail 1982; Staddon 2016; Behrens et al. 2008). Theory has shown that
natural selection favours these associations in complex environments
where conditions are hard to predict (Dridi and Lehmann 2016). Besides
its wide taxonomic and ecological relevance, associative learning is a
flexible cognitive mechanism whose underpinnings show interspecific
variation (Enquist, Lind, and Ghirlanda 2016; Quiñones et al. 2019), so
presumably, it has been modified by natural selection. Despite all this,
associative learning is not often included in the narrative of
evolutionary explanations of behavioural patterns (Fawcett, Hamblin, and
Giraldeau 2013; Kamil 1983; McAuliffe and Thornton 2015). Computational
models of evolution can play an important role to overcome the lack of
integration between learning and evolution. Reinforcement learning
theory encompasses a series of computational methods inspired on the
psychological and neurological mechanisms of associative learning
(Sutton and Barto 2018). These set of algorithms allow the
implementation of biologically realistic problems, capturing the essence
of learning processes (Frankenhuis, Panchanathan, and Barto 2018;
Quiñones et al. 2019). Furthermore, these algorithms can be embedded in
evolutionary simulations to generate theoretical predictions of the
effect that learning can have in behavioural evolution (Leimar and
McNamara 2019).

In here we present an evolutionary model where individuals use
associative learning to develop a tendency to behave aggressively (or
peacefully), in the context of competition over resources, depending on
the quantitative morphological trait they perceive in their opponent
(badge). Over evolutionary time individuals evolve the size of their
badge, and whether it depends on their quality. Under this simple set
up, individuals can use the badge as a signal of quality. We use the
model to assess under what conditions of interaction structure we expect
different communication signals to evolve.

@@

We model the evolution of signals of quality in the context of agonistic
interactions. For simplicity we consider a population of haploid
individuals with non-overlapping generations. Individuals are born every
generation with a level of quality (\(Q_i\), where \(i\) is subscript of
the population vector of size \(N\)) given by a number between zero and
one, which is drawn from a truncated normal distribution
\(N(\mu = 0.5,\sigma)\). An individual with quality \(0\) has the lowest
RHP, while an individual with quality \(1\) has the highest. As part of
development processes individuals produce a morphological signal
(badge), the size of such depends on their level of quality according to
a reaction norm given by \(B_i=1/(1+e^{\alpha_i-\beta_i Q_i})\); where
are \(\alpha_i\) and \(\beta_i\) are individual specific traits that
determine the shape of the reaction norm (Fig. \ref{fig:model-struc} A).
Variation in \(\alpha_i\) and \(\beta_i\) means individuals can have
either uninformative (flat) or informative (logistic) norms. The size of
the signal is constrained between zero and one. We assume different
values of these traits are given by different alleles, and are inherited
from mother to offspring. Unless, with a small probability, mutation
changes the allelic value of the offspring. After birth individuals go
through a round of viability selection. Individual specific survival
probability is given by \(s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})\). Where
\(k_1\) and \(k_2\) are parameter values determining the shape of the
survival function. Importantly, if \(k_2=0\) survival probability is
independent of quality, while if \(k_2>0\) lower quality individuals pay
a higher price for similar-sized badge, fulfilling the core assumption
of the handicap principle (Botero et al. 2010; Grafen 1990; Johnstone
and Norris 1993).

Individuals who survive engage in a series of pairwise interactions
where they compete for resources. Individuals in real populations
typically do not interact at random nor with all individuals in the
population (Kurvers et al. 2014). In order to allow for non-random
interactions, we use the population vector to define an interaction
neighbourhood for the focal individual, the size of the neighbourhood
(\(g\)) is the number of positions around the focal's from which the
partner is drawn. So, if \(g=N\) interactions are global; as \(g\) gets
smaller interactions are more local. In an interaction each individual
must decide whether to escalate a fight or not. Following the classic
\emph{hawk-dove} game (Maynard-Smith 1982), if the focal individual
fights and its partner does not, the focal gets as pay-off the resource
of value \(V\) while its partner gets nothing; if both individuals
restrain from fighting they split up the resource in half; if both
decide to fight they split up the cost of fighting and the winner takes
over the resource. We further assume that the probability of wining a
fight for the focal individual depends on the difference in quality
between her and her interacting partner; specifically it is given by
\(p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})\), where \(k_3\) is a parameter
defining how strong the quality difference determines the winning
probability; and \(i\) and \(j\) denote the the position in the
population vector of the focal and its interacting partner respectively.

The decision of whether to escalate a fight against an interacting
partner can be dependent on the size of the partner's badge, the actual
dependency is determined by the focal´s experiences through a learning
process. We implement learning using the actor-critic approach from
reinforcement learning theory {[}RL; Sutton and Barto (2018);Quiñones et
al. (2019); Leimar and McNamara (2019){]}. Individuals estimate the
reward (pay-off) expected from interacting with partners of different
badge sizes (the critic in RL terminology). After every interaction they
update the estimate of reward proportionally to the difference between
their current estimate and the observed reward (prediction error
\(\delta\)) and to the speed of learning (\(\mathrm{A}\)). Furthermore,
they express different probabilities of retreating (attacking) depending
on the badge size of their opponent (the actor in RL). They update the
probability of retreating (attacking) up or down depending on whether
retreating (attacking) leads to an increase in the reward estimation.
So, if a focal individual decides to escalate a fight against an
individual with small badge, and that leads to an increase in the reward
estimation, the focal individual will increase the probability of
escalating fights with individuals of small badges in the future. Given
that badge size is a real number between 0 and 1, there are infinitely
many badge sizes. Thus, the reward estimation, as well as the
probability of retreating (escalating), must be generalized across
different values. To implement generalization we use the linear function
approximation method based on radial basis functions (Sutton and Barto
2018). Specifically, we pick \(c\) feature centres, which are evenly
spaced values along the badge size interval ({[}0,1{]}). Each one of
these features centres is associated with a weight for reward estimation
and tendency to play retreat in a given interaction. The reward
estimation and the tendency to retreat are calculated (in every
interaction) as the sum of the weights associated with each feature,
weighted by the response triggered by the feature (Fig.
\ref{fig:model-struc} B, dots represent the feature weights) The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:model-struc} B, grey line). Formally the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,\\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-c_z|}{2\theta^2})} 
\end{equation}

where \(x_z\) is the weight of feature centre \(z\) on the reward
estimation; and \(\theta\) is the width of the generalization function.
Similarly, the tendency to retreat is given by the sum of feature
weights associated with the actor, and the probability is obtained by
applying a logistic transformation (Fig. \ref{fig:model-struc} B, black
line). So formally, the log-odds to retreat when facing individual \(j\)
is

\begin{equation}
logit(q_{ij}) =  \sum_{z=1}^{c} y_z e^{(-\frac{|B_i-c_z|}{2\theta^2})} 
\end{equation}

@@

After all their interactions, individuals in the population reproduce
proportionally to their fitness \(w_i\), which is a sum of the baseline
fitness (\(w_0\)) and all the pay-offs obtained throughout their life.
Thus, the combination of natural selection and genetic drift changes the
distribution of values in \(\alpha\) and \(\beta\) that segregates in
the population, effectively changing the badges expressed and the
communication system.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \AddToHook{env/abstract/before}{\vspace*{-4.5cm}}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{unicode-math}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}            
  \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={ The role of learning in the evolution of status signalling: a modeling approach},
  pdfkeywords={My keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\large \textbf{Title:} The role of learning in the evolution of
status signalling: a modeling approach}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\normalsize \textbf{Abbreviated title:} Learning and evolution
in status signalling}
\date{}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\newcommand{\getAff}[1]{
  \getIndex{#1}{}
}

\begin{document}
\maketitle
\begin{flushleft}
\end{flushleft}
\begin{abstract}
Adaptive behavioural responses often depend on qualities of the
interacting partner of an individual. For example, when competing for
resources, an individual might be better off escalating fights with
individuals of lower quality, while restraining from fighting
individuals of higher quality. Communication systems involving signals
of quality allow individuals to reduce uncertainty regarding the
fighting ability of their partners and make more adaptive behavioral
decisions. However, dishonest individuals can destabilize such
communications systems. An open question is whether cognitive
mechanisms, such as learning, can maintain the honesty of signals, thus
favoring their evolutionary stability. We present evolutionary
simulations where individuals can produce a signal proportional to their
quality and learn along their lifetime the best response to the signals
emitted by their peers. Our simulations replicate previous results where
the handicap principle mediates the evolution of signals. In our
simulations learning on the receiver side can mediate the evolution of
signals of quality on the sender side. When the cost of the signal is
proportional to the quality of the sender, all individuals in
populations are honest signalers. In contrast to traditional models
which predict the absence of signals when the cost is not proportional
to the quality of the signaler, our model revealed that learning
facilitates the evolution of a polymorphism in which populations
comprise both honest and dishonest signalers. We argue that learning may
have a role in the evolution and dynamics of a wide range of
communication systems and more generally in behavioral responses.
\end{abstract}
\subsection*{Lay summary}
Biological signals allow individuals to respond appropietly. For
example, a singal of fighting hability would allow peers to decide
whether to fight another individual. But, what prevents a weaklings from
faking to be strong? Using a theoretical approach we assess whether
learning has a role in maintaining signals honest. We find that when
individuals learn how best to respond to a signal, populations are
composed of a mixed of honest and dishonest signals.

\newpage

@@

The outcome of interactions with conspecific individuals is a critical
determinant of fitness in social animals. Irrespective of whether such
interactions are cooperative or competitive, the actions of interacting
individuals influence their reproductive success. However, the best
action for an individual engaging in a social interaction might vary
depending on its own condition and that of individuals with whom it
interacts. Thus, acting based on information about interacting partners
is typically adaptive (Quiñones et al. 2016), yet acquiring such
information is far from trivial. In some cases, interacting partners
(\emph{i.e.} signallers) might be `willing' to provide accurate
information, but in others it might be in their own interest to conceal
information (Johnstone 1997) or to provide deceptive information
(Johnstone and Norris 1993). Typically, in a given context, some
individuals benefit may from broadcasting accurate information, whereas
others may benefit from concealing it. Take, for instance, an
interaction between two individuals where one can help the other. Given
some costs and benefits, a potential donor is likely to be interested in
helping related individuals. Therefore, for a relative of the donor,
broadcasting kinship would be advantageous, whereas concealing the lack
of kinship would be better for an unrelated individual. Similar
scenarios may apply to a variety of aspects of social life such as
finding mates, feeding offspring, or engaging in dominance relationships
or aggressive interactions (Bradbury and Vehrencamp 2011; Møller 1988;
Tibbetts and Dale 2004).

In dominance and aggressive interactions, a crucial piece of information
to guide individual actions is the fighting ability of interacting
partners, which is sometimes referred to as Resource Holding Potential
(RHP, Parker 1974) or simply as quality. Responsiveness to the quality
of an opponent is central to communication systems in which individuals
exhibit badges of status, where a signal typically conveys quality
(Johnstone and Norris 1993; Rohwer 1975). Such signals tend to be
arbitrary, however, meaning the signals could potentially be produced by
high- and low-quality individuals alike, opening up the possibility of
dishonest signalling and raising the question of evolutionary stability.
Signals can be evolutionarily stable whenever low-quality individuals
pay a higher fitness cost for carrying the signal such that it is not in
the interest of low-quality individuals to fake quality by producing a
large badge. Signal costs inversely proportional to quality may come
about as a result of the production costs of the signal, in which case
the signal works like a handicap (Botero et al. 2010; Johnstone and
Norris 1993; Grafen 1990; Zahavi 1975). Alternatively, the costs can
come from the aggressive reaction of receivers when the convention of
the communication is broken (Enquist, Ghirlanda, and Hurd 2010; Tibbetts
and Dale 2004). For example in paper wasps, subordinate individuals with
experimentally manipulated dominant-like facial patterns received more
aggression from the dominant (Tibbetts and Dale 2004). In this case,
costs are socially rather than developmentally mediated and are
triggered by the violation of the convention, although how the
convention is established in the first place is unclear.

A somewhat ignored component of communication systems in the context of
aggressive interactions are cognitive aspects of the receiver module.
Theoretical models often assume that communication systems relying on
badges of status have reaction norms as their mechanistic underpinning
(Botero et al. 2010), with individuals using their own quality and the
badges of opponents to determine whether to aggressively engage in
contests. Reaction norms allow individuals to respond to the available
information without a large cognitive burden because the computations
involved in the reaction norm response do not require much memory nor
advanced algorithmic processes. This contrasts with other
information-processing mechanisms like individual recognition where
individuals associate cues of their peers to their quality. Because
these associations must be learnt throughout the life of individuals,
they have the usual cognitive requirements of associative learning
processes. An alternative view of badges of status is that individuals
learn to react to them based on their experiences (Guilford and Dawkins
1991), which would imply that receivers must learn to associate signals
with the fighting ability of bearers just as in systems based on
individual recognition. In cases where badges of status vary
quantitatively (\emph{e.g.} in size or intensity), fighting ability may
increase monotonically with attributes of the badge and would be
reinforced by every interaction. Therefore, in principle, learning in
such scenarios would be faster than in systems involving individual
recognition in which the association between signals and their meaning
varies depending on the interacting partners. In any case, learning may
be a central cognitive mechanism in both types of communication systems,
but the role of learning in these contexts has not been thoroughly
explored in the empirical nor theoretical literature.

Associative learning is a taxonomically widespread cognitive mechanism
that allows individuals to associate rewards with environmental stimuli
and thus behave adaptively (Heyes 2012; Macphail 1982; Staddon 2016;
Behrens et al. 2008). Theory has shown that natural selection favours
such associations in complex environments where conditions are difficult
to predict (Dridi and Lehmann 2016). Associative learning is a flexible
cognitive mechanism whose underpinnings show interspecific variation
(Enquist, Lind, and Ghirlanda 2016; Quiñones et al. 2020; Prétôt,
Bshary, and Brosnan 2016; Prétôt et al. 2021). For example, species can
have different learning rates (Hoedjes et al. 2010), some may vary these
rates during learning trials which allows them to have more flexible
learning (Leimar, Quiñones, and Bshary 2023), and some may include
future rewards in their associations (Raby et al. 2007; Osvath 2009).
Associative learning is not a single mechanism, but rather a set of
cognitive structures and processes that can vary in their scope and
complexity. Presumably, these structures and processes have been
modified by natural selection and could provide novel explanations for
behavioural variation. Nonetheless, associative learning is not often
included in the narrative of evolutionary explanations of behavioural
patterns (Fawcett, Hamblin, and Giraldeau 2013; Kamil 1983; McAuliffe
and Thornton 2015).

Computational models of evolution can overcome the lack of integration
between learning and evolution. Reinforcement learning theory
encompasses a series of computational methods inspired on the
psychological and neurological mechanisms of associative learning
(Sutton and Barto 2018). This set of algorithms allows the
implementation of biologically realistic problems, capturing the essence
of learning processes (Frankenhuis, Panchanathan, and Barto 2018;
Quiñones et al. 2020). Furthermore, these algorithms can be embedded in
evolutionary simulations to generate theoretical predictions of the
effect of learning on behavioural evolution (Leimar and McNamara 2019;
Leimar and Bshary 2022b).

Here, we present an evolutionary model where individuals use associative
learning to develop a tendency to behave either aggressively or
peacefully in the context of competition over resources, depending on a
quantitative morphological trait (\emph{i.e.} a badge indicating
quality) they perceive in their opponents. Over evolutionary time, the
size of the badge evolves as does its dependency on the individual's
quality. Under this simple set up, individuals can use the badge as a
signal of quality. We use the model to assess under what conditions one
might expect communication signals to evolve as handicaps or
conventions.

@@

We model the evolution of signals indicating individual quality in the
context of agonistic interactions. For simplicity, we consider a
population of haploid individuals with non-overlapping generations.
Individuals are born every generation with a level of quality (\(Q_i\),
where \(i\) is subscript of the population vector of size \(N\)) given
by a number between zero and one, which is drawn from a truncated normal
distribution \(N( 0.5,\sigma)\). An individual with quality \(0\) has
the lowest RHP, while an individual with quality \(1\) has the highest.
As they develop, individuals produce a phenotypic signal (badge), the
size of which depends on their quality according to a reaction norm
given by \(B_i=1/(1+e^{\alpha_i-\beta_i Q_i})\); where \(\alpha_i\) and
\(\beta_i\) are individual specific traits that determine the shape of
the reaction norm (Fig. \ref{fig:modelStruc} A). Variation in
\(\alpha_i\) and \(\beta_i\) means individuals can have either
uninformative (flat) or informative (logistic) norms. Informative
reaction norms represent a developmental program where the phenotype of
the individual is determined by the environmental conditions under which
it grows, thus causing a correlation between quality and signal. The
size of the signal is constrained to take values between zero and one.
We assume different values of these traits are given by different
alleles and are inherited from mother to offspring unless, with a small
probability (\(\mu\)), mutation changes the allelic value of the
offspring by an amount drawn from a normal distribution
\(TN(0,\sigma_\mu)\).

After birth, individuals go through a round of viability selection. The
survival probability of an individual is given by
\(s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})\), where \(k_1\) and \(k_2\) are
parameter values determining the shape of the survival function.
Importantly, if \(k_2=0\) survival probability is independent of
quality, while if \(k_2>0\) lower-quality individuals pay a higher price
for similar-sized badges, fulfilling the core assumption of the handicap
principle (Botero et al. 2010; Grafen 1990; Johnstone and Norris 1993).

Individuals who survive engage in a series of pairwise interactions
where they compete for resources. In each interaction individuals must
decide whether to escalate a fight or not. Following the classic
\emph{hawk-dove} game (Maynard-Smith 1982), if the focal individual
fights and its partner does not, then the focal gets as pay-off the
resource of value \(V\) while its partner gets nothing; if both
individuals restrain from fighting, then they split up the resource in
half. If both individuals decide to fight, then the winner takes over
the resource and the cost of fighting is split between the two players.
We further assume that the probability of wining a fight for the focal
individual depends on the difference in quality between it and its
interacting partner; specifically it is given by
\(p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})\), where \(k_3\) is a parameter
defining how strong the quality difference determines the winning
probability, and \(i\) and \(j\) denote the the position in the
population vector of the focal and its interacting partner,
respectively.

The decision of whether to escalate a fight against an interacting
partner can depend on the size of the partner's badge, with the
dependency being determined by the focal's experiences acquired through
a learning process. We implement learning using the actor-critic
approach from reinforcement learning (RL) theory (Sutton and Barto 2018;
Quiñones et al. 2020; Leimar and McNamara 2019; Leimar 2021).
Individuals estimate the reward (pay-off) expected from interacting with
partners of different badge sizes (the critic in RL terminology). After
each interaction they update the estimate of reward proportionally to
the difference between their current estimate and the observed reward
(prediction error \(\delta\)) and to the speed of learning
(\(\mathrm{A}\)). Furthermore, individuals express different
probabilities of retreating/attacking depending on the badge size of
their opponent (the actor in RL). They update the probability of
retreating/attacking depending on whether retreating/attacking leads to
an increase in the reward estimation. Thus, if a focal individual
decides to escalate a fight against an individual with a small badge and
this leads to an increase in the reward estimation, then the focal
individual will increase the probability of escalating fights with
individuals of small badges in the future. Given that badge size is a
real number between 0 and 1, there are infinitely many badge sizes.
Thus, the reward estimation, as well as the probability of
retreating/attacking, must be generalized across different values. To
implement generalization we use the linear function approximation method
based on radial basis functions (Sutton and Barto 2018). Specifically,
we pick \(c\) feature centres, which are evenly spaced values along the
badge size interval ({[}0,1{]}) where the updates are focused. For
simplicity, we keep the location of these feature centres constant and
the same for all individuals; they are stored in vector \(\symbf{b}\) .
Each of these feature centres is associated with a weight for reward
estimation and tendency to play retreat in a given interaction. The
reward estimation and the tendency to retreat are calculated (in every
interaction) as the sum of the weights associated with each feature
centre, weighted by the response triggered by the feature (Fig.
\ref{fig:modelStruc} B, dots represent the feature weights). The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,\\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

where \(x_z\) is the weight of feature centre \(z\) on the reward
estimation; and \(\theta\) is the width of the generalization function.
Similarly, the tendency to retreat is given by the sum of feature
weights associated with the actor, and the probability is obtained by
applying a logistic transformation (Fig. \ref{fig:modelStruc} B, black
line). Thus formally, the log-odds to retreat when facing individual
\(j\) is

\begin{equation}
logit(q_{ij}) =  \sum_{z=1}^{c} y_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual \(n\) times and the
interaction partner is chosen at random. Thus, the expected number of
interactions for each individual is \(n+\frac{n(N+1)}{N}\). After the
interaction round , individuals in the population reproduce with a
probability proportional to their total pay-off \(w_i\), which is a sum
of the baseline pay-off (\(w_0\)) and all the pay-offs obtained
throughout their life. Thus, the combination of natural selection and
genetic drift changes the distribution of values in \(\alpha\) and
\(\beta\) that segregates in the population, effectively changing the
badges expressed and the communication system.
